Does It Scale? Static Output Approximation of PHP Web Applications Using Symbolic Execution

### 2017-01-09 paper.tex
update
refactored sec 3

\subsection{Existing Tool Support for PHP}
All tools proposed so far leverage a representation of all possible HTML client
page output. Since any variant can be analyzed in particular, this enables tool
support addressing the web application as a whole in spite of its staged
nature, where parts of the web application (client-side output) are generated
dynamically.

@@

All three tools though have only been applied to systems that are small and not
maintained any more. Despite the functional benefits these tools provide to
developers, their practicality for large and modern systems has not been
studied.

%\subsection{Assumptions and Limitations}
%Symbolic execution remains an analysis technique not dedicated to a particular programming language. For the implementation of the symbolic execution engine in \cite{Nguyen:2014:BCG:2635868.2635928,Nguyen:2015:CPS:2786805.2786872}, several assumptions have been made to adapt to the PHP language. The following symbolic execution semantics for PHP features differ from concrete execution since symbolic values need to be handled without ambiguity.

%First, loops in the previous execution semantics are only executed once instead of an arbitrary number of times. Since a loop condition can contain symbolic values, it might not be feasible to determine when a loop terminates. Hence, any output constructed during the single loop iteration is highlighted to possibly be repeated  an arbitrary number of times. Second, recursive functions are also only executed once per call, and any subsequent call ot that function returns a symbolic value. This, again, is due to the difficulty to determine the exact recursion depth similar to the number of loop iterations. Finally, since the symbolic execution engine applies a static analysis, any form of user interaction providing input data as well as any form of input referred to from the deployment context of the web application, i.e. configuration files or databases, is substituted by a symbolic value.

%Language-wise the symbolic execution semantics for the tools presented in Section \ref{sec:workflow} currently does not support object-oriented programming. PHP provides a large number of built-in functions to manipulate strings as well as arrays. The current symbolic execution engine only provides implementations for a small range of PHP standard library functions, mainly string operations.

\section{Scalable Output Approximation}%
The main objective of this experience report is to investigate whether the
tools described in Section \ref{sec:stateoftheart} are practical. In particular
we want to find out whether symbolic execution as the underlying technique for
static output approximation scales for PHP, and what may limitations restrict
scalability. We may divide possible limitations into either matters of
engineering or conceptual limitations where no precise execution is possible.

In the following we describe in detail our implementation procedure, practical
problems we encountered and whether these are limiting scalability conceptually
or just require en gineering effort to be tamed.

\subsection{Towards Scalable Output Approximation?}
We re-implement\footnote{\url{https://github.com/smba/oak}} the existing
symbolic execution semantics of \cite{Nguyen:2014:BCG:2635868.2635928} with the
goal to support modern large-scale dynamic web applications, or at least to
better understand possible limitations. For reference, we choose to aim to
support  \textsf{WordPress}
\footnote{\url{https://github.com/WordPress/WordPress }}, a popular Content
Management System of reasonable size that is still maintained. Moreover, our
implementation uses parts of
Quercus\footnote{\url{http://quercus.caucho.com/}}, a pure Java implementation
of the PHP interpreter, which is tested against the \textsf{WordPress} test
suite. We choose a test-driven approach for our implementation with regression
tests and continuous integration. Our test cases are based on both custom
examples for different language features as well as \textsf{SchoolMate}, a 
small-scale system, and, at least in parts, \textsf{WordPress}, whose output we
manually validate and then automatically re-check for each test run.

Our implementation basically follows the symbolic execution semantics of
\cite{Nguyen:2014:BCG:2635868.2635928}, yet we extend it with additional rules
for class definitions, class instantiation, and method invocation to support
object-oriented programming. Furthermore, we add rules for do-while and 
foreach loops since these are not supported as yet. Note that we execute all
iterations of a foreach loop since the number of iterations is known since we
 are traversing all items of a given array.

@@

\subsection{Experiment Setup} 
For the evaluation we symbolically executed a selection of PHP systems. We
selected a corpus of twelve PHP systems with regard to system size and recency
as the corpus of the case study for \cite{Nguyen:2014:BCG:2635868.2635928} only
contained small-scale systems that are not maintained any more. The full list
of PHP systems is shown in Table \ref{corpus}. Our selection of PHP systems
includes

>>>>

\subsection{Existing Tool Support for PHP}\label{sec:existing_tools}
All tools proposed so far leverage a representation of all possible HTML client
page output. Since any variant can be analyzed in particular, this enables tool
support addressing the web application as a whole in spite of its staged
nature, where parts of the web application (client-side output) are generated
dynamically.

@@

\subsection{Assumptions And Limitations}\label{sec:limitations}
In addition to the limitations introduced by  symbolic execution, such as symbolic inputs and one-time loop execution, the symbolic execution described in \cite{Nguyen:2014:BCG:2635868.2635928} does not support advanced language features, including object-oriented programming. Moreover, the range of standard library functions supported are mostly string operations, yet PHP provides a large collection of functions for array operations.
All three tools described in Section \ref{sec:existing_tools} though have only been applied to systems that are small and not maintained any more. Despite the functional benefits these tools provide to developers, their practicality for large and modern systems with respect to advanced language support has not been evaluated.

\section{Scalable Output Approximation}%
The main objective of this experience report is to investigate whether we can have built practical tools that are on output approximations like the tools in Section \ref{sec:existing_tools}. In particular, we want to find out whether symbolic execution as the underlying technique for static output approximation is scalable, i.e., it computes an accurate approximation for large and modern PHP systems with reasonable time and space consumption.

Also, we are interested in the limitations of scalable symbolic execution for PHP web applications. Limitations may either be conceptional, where symbolic execution of PHP code, in contrast to concrete execution, is inaccurate: For example an include expression evaluated to a symbolic value simply provides to little information to include a script. Or, in spite of infeasible effort required, limitations could allow an accurate symbolic execution. For example, exhaustively exploring all feasible paths in an application may result in an infeasible number of program states to store; nevertheless, this so called state explosion problem may be addressed with additional engineering.

\subsection{Towards Scalable Output Approximation?}
As our goal is to have practical tools support, we approach this question by re-implementing the symbolic execution semantics described in \cite{Nguyen:2014:BCG:2635868.2635928} and extending it with support for additional language features. We choose to aim to support symbolic execution for a real-world example system of reasonable size that incorporates the missing language features described in Section \ref{sec:limitations} and is still developed and maintained: \textsf{WordPress} is a popular open-source Content Management System (CMS) with around 300k lines that provides a vivid plug-in environment and is maintained by a large community. 
For our implementation of a symbolic execution engine for PHP, \textsf{Oak}, we have chosen a test-driven approach with continuous integration using regression tests based on \textsf{WordPress} and \textsf{SchoolMate} (see Section \ref{sec:experiment_setup}). Oak incorporates the existing execution semantics of \cite{Nguyen:2014:BCG:2635868.2635928} and extends it with support for object-oriented orientation (class definition, class instantiation and method invocation) as well as PHP array operations. A list of the functions provided by the PHP standard library can be found at \cite{PHP}.
Throughout the implementation we did not achieve high code coverage for WordPress, which was the main motivation to further investigate limitations in the wild.

\subsection{What Went Wrong? Experience Report}

@@

\subsection{Experiment Setup} \label{sec:experiment_setup}
For the evaluation we symbolically executed a selection of PHP systems. We
selected a corpus of twelve PHP systems with regard to system size and recency
as the corpus of the case study for \cite{Nguyen:2014:BCG:2635868.2635928} only
contained small-scale systems that are not maintained any more. The full list
of PHP systems is shown in Table \ref{corpus}. Our selection of PHP systems
includes
